---
title: "Cap DAP"
format: html
editor: visual
---

Setup:

```{r}
rm(list=ls()) # cleans the environment
library(tidyverse) # imports the tidyverse package
library(here) # imports the here package
library(ggfortify) # imports the ggfortify package
```

Import the data set:

```{r}
diabetes <- read.csv(here("data", "BMI and Diabetes.csv")) 
# import the data and assign it to the variable name: "diabetes"
```

Check for mistakes and outliers

```{r}
ggplot(diabetes, aes(x=Cholesterol, y=BMI)) +
  geom_point()
```

Look at data:

```{r}
glimpse(diabetes)
```

Turn gender into a factor and turn 1s into male and 2s into female:

```{r}
fix_gender_M <- which(diabetes$Gender == 1)
diabetes$Gender[fix_gender_M] <- "Male" # rename all 1 factors into male

fix_gender_F <- which(diabetes$Gender == 2)
diabetes$Gender[fix_gender_F] <- "Female" #rename all 2 factors into female

diabetes$Gender <- factor(diabetes$Gender) # reassign the values
```

Turn family_history, drinking_status, smoking_status, and site into factors
```{r}
diabetes$family_history <- as.factor(diabetes$family_history)

diabetes$drinking_status <- as.factor(diabetes$drinking_status)

diabetes$smoking_status <- as.factor(diabetes$smoking_status)

diabetes$site <- as.factor(diabetes$site)
```

create a mean arterial pressure column:
```{r}
diabetes <- diabetes |> mutate(MAP = (DBP + (1/3)*(SBP-DBP)))
```

create a BMI_cat column:
```{r}
diabetes <- diabetes |> 
  mutate(BMI_cat = case_when(BMI < 18.5 ~ "Underweight",
                             BMI >= 18.5 & BMI < 25 ~ "Healthy",
                             BMI >= 25 & BMI < 30 ~ "Overweight",
                             BMI >= 30 ~ "Obese"))
```


Exploratory data analysis:

A) plot BMI vs diabetes diagnosed and interpret the results:
```{r}
ggplot(diabetes, aes(BMI)) +
  geom_bar(width=1) +
  facet_wrap(diabetes$diabetes_diagnosed, scales = "free_y") # this allows us to see the different graphs more clearly despite having fewer samples for when diabetes_diagnosed = 1
```
The average mean for diabetes diagnosed appears to be around 28 BMI while the average mean for no diabetes diagnosed appears to be around 22. I think that there will be a significant difference in the mean BMI for diagnosis of diabetes. 

B) create a linear model
```{r}
diagnosed_model <- lm(BMI ~ diabetes_diagnosed, data = diabetes)
```

C) check the assumptions
```{r}
autoplot(diagnosed_model, smooth.colour = NA)
```


Mean BMI and SD for diabetes diagnosed and not diagnosed:

```{r}
diabetes_bmi <- diabetes |> 
  group_by(diabetes_diagnosed) |>
  summarise(mean_bmi = mean(BMI), 
            sd_bmi = sd(BMI))
```




BMI by blood pressure:

a) graph BMI vs MAP and interpret the results
```{r}
ggplot(diabetes, aes(x = MAP, y = BMI)) + 
  geom_point()
```
There may be a slight positive association between MAP and BMI.

b) create a model
```{r}
MAP_model <- lm(MAP ~ BMI, data = diabetes)
```

c) check the assumptions
```{r}
autoplot(MAP_model, smooth.colour=NA)
```

d) create the regression test:
```{r}
summary(MAP_model)
```

e) create a final graph
```{r}
ggplot(diabetes, aes(x=MAP, y=BMI)) +
  geom_point()
```



a) graph triglycerides vs MAP and interpret
```{r}
ggplot(diabetes, aes(x = MAP, y = Triglycerides)) + 
  geom_point()
```
