---
title: "Braun-Preliminary-Analysis-Q3"
format: html
editor: visual
---

In this file, I will be using the diabetes.csv data set to analyze BMI categories and FPG The question I will be asking is:

Is there a difference in mean FPG between individuals the different categories of BMI?

I will analyze this question using an ANOVA test because I have a ratio response variable (FPG), a categorical predictor variable (BMI_cat), I am looking to see if there is a difference between group means, samples are independent because the FPG of each person is not related to another persons FPG, and I have 4 groups to compare (underweight, healthy, overweight, obese).

My null hypothesis is that there will be no difference in mean FPG between different BMI categories.

## Setup

```{r}
rm(list=ls()) # clear environment
library(tidyverse) # import tidyverse package
library(here) # import here package 
library(ggfortify) # import ggfortify package
library(multcomp) # import multcomp package
```

## Import the data

```{r}
diabetes <- read.csv(here("data", "diabetes.csv"))
# this is the cleaned version of my original dataset
```

We need to turn BMI_cat into a factor variable:

```{r}
diabetes$BMI_cat <- as.factor(diabetes$BMI_cat)
# turn BMI_Cat into a factor with four levels (underweight, healthy, overweight, obese)
```

## Initial Plot and Interpretation:

Plot a boxplot with BMI_cat and FPG

```{r}
ggplot(diabetes, aes(x=BMI_cat, y=FPG)) +
  geom_boxplot()
# creates a boxplot of FPG for each BMI category
```

The mean FPG for healthy is around 4.5, 5 for Obese, 4.7 for overweight, and 4.5 for underweight. I believe that there will be a significant difference in mean FPG in at least two groups.

## Create a model

```{r}
FPG_model <- lm(FPG ~ BMI_cat, data = diabetes)
# use `lm()` to build a linear model with BMI as the response and MAP as predictor
```

FPG acts as the response variable and BMI_cat as the predictor variable. The linear model will allow us to run an ANOVA test.

## Check assumptions

```{r}
autoplot(FPG_model)
```

The first plot shows us the residuals vs fitted which tells us if our data is a linear model. The blue line is straight which is good.

The normal Q-Q plot shows us if our data meets the requirements for normality. Most of the points fall on or close to the line which is good.

The Scale-Location plot shows us if our data meets the requirements for equal variance. There is no real pattern to the data so it looks good.

The residuals vs levels plot helps us see if there are any points that have a lot of leverage and can effect the line. The line is straight which is good.

## Run ANOVA

```{r}
anova(FPG_model)
```

The anova table shows that there is a significant difference in mean FPG between the groups so we can reject our null hypothesis.

```{r}
summary(FPG_model)
```

The intercept is the healthy category because it comes first alphabetically. The summary shows how different each category is from healthy. Obese has a significantly higher mean FPG than healthy, overweight has a significantly higher FPG than healthy, and underweight has a significantly lower FPG than healthy. (p-value \<\< .001)

## Run tukey test

```{r}
phc1 <- glht(FPG_model, linfct = mcp(BMI_cat="Tukey"))
summary(phc1)
```

The tukey test shows that: - Obese has a significantly higher mean FPG than healthy (p \<\< .001) - Overweight has a significantly higher mean FPG than healthy (p \<\< .001) - Underweight has a significantly lower mean FPG than healthy (p \<\< .001) - Overweight has a significantly lower mean FPG than obese (p \<\< .001) - Underweight has a significantly lower mean FPG than obese (p \<\< .001) - Underweight has a significantly lower mean FPG than overweight (p \<\< .001)

## Show the compact letter display

```{r}
cld(phc1) # different letters signify a significant difference
```

Each BMI category has a different letter which means they are all significantly different from each other.

## Make a better plot

```{r}
ggplot(diabetes, aes(x=BMI_cat, y=FPG)) +
  geom_boxplot(aes(fill = BMI_cat)) +
  labs(x="BMI Category", y="FPG (mmHg)") +
  coord_flip() +
  annotate("text", x = c(4.2,3.2,2.2,1.2), y=c(5.5,5.8,6,5.6),
           label=c("A","B", "C", "D"))
# coloring the boxes by their BMI category, flipping the coordinates so it is easier to read category names, label x and y axis, and add the CLD letters to show where significant differences lie
```

This plot shows the boxplots of FPG for each BMI category. Different letters signify significant differences.
