---
title: "Final Analysis"
format: html
editor: visual
---

In this file, I will be using the diabetes.csv data set to analyze BMI and MAP. The question I will be asking is:

Is there a relationship between Body Mass Index (BMI) and Mean Arterial Pressure (MAP)?

I will analyze this question using a regression test because I have a ratio response variable (BMI), a continuous predictor variable (MAP), I am looking to see if there is a relationship between two variables, and I only have 1 predictor variable.

My null hypothesis is that there will be no relationship between BMI people and MAP.

## Setup

```{r}
rm(list=ls()) # clear environment
library(tidyverse) # import tidyverse package
library(here) # import here package 
library(ggfortify) # import ggfortify package
```

## Import the data

```{r}
diabetes <- read.csv(here("data", "diabetes.csv"))
# this is the cleaned version of my original data set
```

## Initial Plot and Interpretation:

```{r}
ggplot(diabetes, aes(x=MAP, y=BMI)) +
  geom_point()
# create a dotplot to see if there appears to be any relationship (positive, negative, linear, etc.)
```

There appears to be a positive relationship between BMI and MAP. The BMI appears to range between 14 and 53. The MAP appears to range between 45 and 172.

The slope can be calculated using rise/run so 53-14 = 39 and 172-45 = 127.

So the slope = 39/127 = 0.307.

We can estimate the y-intercept to be around 20.

## Create the model:

```{r}
MAP_model <- lm(BMI ~ MAP, data = diabetes)
# use `lm()` to build a linear model with BMI as the response and MAP as predictor

```

BMI acts as the response variable and MAP as the predictor variable. The linear model will allow us to run a regression analysis.

## Check the assumptions:

```{r}
autoplot(MAP_model)
```

The first plot shows us the residuals vs fitted which tells us if our data is a linear model. The blue line curves down slightly at the end but it is mostly straight which is good.

The normal Q-Q plot shows us if our data meets the requirements for normality. Most of the points fall on or close to the line. The points increases towards the ends of the line but it still looks pretty good.

The Scale-Location plot shows us if our data meets the requirements for equal variance. There is no real pattern to the data so it looks good. The blue line is curved down a bit at the beginning but it looks like it should be ok.

The residuals vs levels plot helps us see if there are any points that have a lot of leverage and can effect the line. The blue line curves down quite a bit at the end but it should be ok.

```{r}
anova(MAP_model)
```

The anova table shows us that our null hypothesis can be rejected and that there is a relationship between BMI and MAP. The p-value \<\< .001 which indicates that the slope is significantly different from 0.

```{r}
summary(MAP_model)
```

This output tells us that the slope is 0.168 and the y-intercept is 13.72. It also shows us that the R-squared is 0.1398 which means MAP accounts for about 13.98% of the variation of BMI which is not too much.

## Interpret

We can reject our null hypothesis because the p-value is less than 0.001. This means that there is a relationship between MAP and BMI and that the slope of 0.168 is significantly different from zero. (F_1, 211807 = 34420, p-value \<\< 0.001)

## Make a better plot

```{r}
ggplot(diabetes, aes(x=MAP, y=BMI))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x = "MAP (mmHg)", y = "BMI")+
  theme_bw()
# this plot adds the slope line to the dotplot 
```

This plot displays the slope (0.168) which shows that it is not zero.
